# Exponential Smoothing

We will now get into our first modeling, which are exponential smoothing models.  They have been around for quite some time, but can still be useful if interested in one-step ahead forecasting.

```{r, message=FALSE,echo=FALSE}
library(readr)
library(fpp3)
library(fable)
library(fabletools)
library(lubridate)
library(expsmooth)
library(lmtest)
library(zoo)
library(seasonal)
library(ggplot2)
library(seasonalview)
library(aTSA)

library(reticulate)

use_python("C:\\ProgramData\\Anaconda3\\envs\\R_Env2\\python.exe")
#use_python("C:\\ProgramData\\Anaconda3\\python.exe")

file.dir = "https://raw.githubusercontent.com/sjsimmo2/TimeSeries/master/" 
input.file1 = "usairlines.csv"
input.file2 = "steel.csv"
input.file3 = "leadyear.csv"
input.file4 = "ebay9899.csv"
input.file5 = "fpp_insurance.csv" 
input.file6 = "ar2.csv"
input.file7 = "MA2.csv"
input.file8 = "hurrican.csv"

# Reads the data at specified directory
# If the file directory is incorrect, then this won't run
USAirlines = read_csv(paste(file.dir, input.file1,sep = ""),show_col_types = FALSE)
Steel = read_csv(paste(file.dir, input.file2, sep = ""),show_col_types = FALSE)
Lead.Year = read_csv(paste(file.dir, input.file3, sep = ""),show_col_types = FALSE)
Ebay = read_csv(paste(file.dir, input.file4, sep = ""),show_col_types = FALSE)
Quotes= read_csv(paste(file.dir, input.file5, sep = ""),show_col_types = FALSE)
Y= read_csv(paste(file.dir, input.file6, sep = ""),show_col_types = FALSE)
x=read_csv(paste(file.dir, input.file7, sep = ""),show_col_types = FALSE)
hurricane=read_csv(paste(file.dir, input.file8, sep = ""),show_col_types = FALSE)
temp<-read_csv("Q:\\My Drive\\Fall 2017 - Time Series\\DataR\\minntemp.csv", show_col_types = FALSE)


```

In R, we are able to do Simple (or Single) Exponential Smoothing Models, Holt Exponential Smoothing Models and Holt-Winters Exponential Smoothing Models.  We will be using the Steel data set and the airline data set to illustrate these models.  Each of these are shown below.  

## Simple Exponential Smoothing

For Simple Exponential Smoothing Models (SES), we have only one component, referred to as the level component.
$$\hat{Y}_{t+1}= L_{t}\\
  L_{t} = \alpha Y_{t} + (1-\alpha)L_{t-1}$$
  
This is basically a weighted average with the last observation and the last predicted value.  Since this only has a level component, forecasts from SES models will be a horizontal line (hence why this method is called "one-step ahead" forecasting).  

Before modeling, be sure to divide your data into a training, validation and test (or at least training and test).  The below code illustrates the Simple (Single) Exponential Smoothing Model.

```{r ses}

# Building a Single Exponential Smoothing (SES) Model - Steel Data #
Steel <- Steel |> mutate(date = seq(ymd('1932-07-01'),ymd('1980-01-01'),by='6 months'))

steel_ts<-Steel |> mutate(Month=yearmonth(date)) |> as_tsibble(index=Month)

steel_train <-steel_ts |> filter(year(date) <= 1975)

SES.Steel <- steel_train |>
  model(ETS(steelshp ~ error("A") + trend("N") + season("N")))
Steel.for <- SES.Steel |>
  fabletools::forecast(h = 9)
report(SES.Steel)

# Plot the SES model on steel data
Steel.for |>
  autoplot(steel_ts) +
  geom_line(aes(y = .fitted), col="#D55E00",
            data = augment(SES.Steel)) +
  labs(y="Steel Exports", title="Steel Exports") +
  guides(colour = "none")

# To get fitted values for training data set:
Steel_fitted <-fitted(SES.Steel)$.fitted
# To get fitted values for test data set:
Steel_test <- Steel.for$.mean

# Computes accuracy statistics for SES model on steel data (test data)
fabletools::accuracy(Steel.for, steel_ts)

```

## Holt ESM

The Holt model incorporates trend information.  So, now there are two components: level and trend.  For each component, there will be a smoothing coefficient (or weight).  CAREFUL, when you look at parameter estimates, these are NOT the estimates for the mean nor the linear trend...you should be thinking of them as weights (between 0 and 1).  The overall form for Holt's method is:

$$\hat{Y}_{t+h}= L_{t}+hT_{t}\\
  L_{t} = \alpha Y_{t} + (1-\alpha)(L_{t-1}+T_{t-1})\\
  T_{t} = \beta (L_{t}-L_{t-1}) + (1-\beta) T_{t-1}$$
  
For the Holt's method, when you forecast, you will see a trending line.  

```{r Holt}
# Building a Linear Exponential Smoothing Model - US Airlines Data #

USAirlines_ts <- USAirlines |> mutate(date=myd(paste(Month, Year, "1"))) |> mutate(Month2=yearmonth(date)) |> as_tsibble(index=Month2)

air_train <-USAirlines_ts |> filter(year(date) <= 2005)

LES.air <- air_train |>
  model(ETS(Passengers ~ error("A") + trend("A") + season("N")))
air.for <- LES.air |>
  fabletools::forecast(h = 27)
report(LES.air)

# Plot the data
air.for |>
  autoplot(USAirlines_ts) +
  geom_line(aes(y = .fitted), col="#D55E00",
            data = augment(LES.air)) +
  labs(y="Passengers", title="US Airline Passengers") +
  guides(colour = "none")

```

We can also perform Holt's method with a damped trend.  You will see the formula for the damped trend is similar to the previous Holt formula with an addition of a dampening parameter.

$$\hat{Y}_{t+h}= L_{t}+\sum_{i}^{k}\phi^{i}T_{t}\\
  L_{t} = \alpha Y_{t} + (1-\alpha)(L_{t-1}+\phi T_{t-1})\\
  T_{t} = \beta (L_{t}-L_{t-1}) + (1-\beta) \phi T_{t-1}$$

We will illustrate the damped trend on the Airline data set.

```{r damped trend}
LdES.air <- air_train |>
  model(ETS(Passengers ~ error("A") + trend("Ad") + season("N")))
air.for <- LdES.air |>
  fabletools::forecast(h = 27)
report(LdES.air)

# Plot the data
air.for |>
  autoplot(USAirlines_ts) +
  geom_line(aes(y = .fitted), col="#D55E00",
            data = augment(LdES.air)) +
  labs(y="Passengers", title="US Airline Passengers") +
  guides(colour = "none")

```

## Holt-Winters

The Holt-Winters (HW) model has three components to it (level, trend and seasonality).  Seasonality is an interesting component to model since we can have an additive seasonal component or a multiplicative seasonal component.  Both models are shown below:

Additive HW
$$\hat{Y}_{t+h}= L_{t}+hT_{t} + S_{t-p+h}\\
  L_{t} = \alpha (Y_{t} - S_{t-p}) + (1-\alpha)(L_{t-1}+T_{t-1})\\
  T_{t} = \beta (L_{t}-L_{t-1}) + (1-\beta) T_{t-1}\\
  S_{t} = \gamma (Y_{t}-L_{t-1}-T_{t-1}) + (1-\gamma) S_{t-p}$$

Multiplicative HW
$$\hat{Y}_{t+h}= (L_{t}+hT_{t}) S_{t-p+h}\\
  L_{t} = \alpha \frac{Y_{t}} {S_{t-p}} + (1-\alpha)(L_{t-1}+T_{t-1})\\
  T_{t} = \beta (L_{t}-L_{t-1}) + (1-\beta) T_{t-1}\\
  S_{t} = \gamma \frac{Y_{t}}{L_{t-1}+T_{t-1}} + (1-\gamma) S_{t-p}$$

Where p is the frequency of the seasonality (i.e. how many "seasons" there are within one year).  


```{r Seasonal models}
# Building a Holt-Winters ESM - US Airlines Data - Additive Seasonality

HWadd.air <- air_train |>
  model(ETS(Passengers ~ error("A") + trend("A") + season("A")))
air.for <- HWadd.air |>
  fabletools::forecast(h = 27)
report(HWadd.air)

# Plot the data
air.for |>
  autoplot(USAirlines_ts) +
  geom_line(aes(y = .fitted), col="#D55E00",
            data = augment(HWadd.air)) +
  labs(y="Passengers", title="US Airline Passengers") +
  guides(colour = "none")


### Multiplicative model
HWmult.air <- air_train |>
  model(ETS(Passengers ~ error("M") + trend("A") + season("M")))
air.for <- HWmult.air |>
  fabletools::forecast(h = 27)
report(HWmult.air)

# Plot the data
air.for |>
  autoplot(USAirlines_ts) +
  geom_line(aes(y = .fitted), col="#D55E00",
            data = augment(HWmult.air)) +
  labs(y="Passengers", title="US Airline Passengers") +
  guides(colour = "none")




```

## Comparing forecasts

This should be done on the validation data set (test data should ONLY be used ONCE...at the very end).

```{r forecasts}
air_fit <- air_train |>
  model(
    SES = ETS(Passengers ~ error("A") + trend("N") + season("N")),
    `Linear` = ETS(Passengers ~ error("A") + trend("A") + season("N")),
    `Damped Linear` = ETS(Passengers ~ error("A") + trend("Ad") + season("N")),
    HWAdd = ETS(Passengers ~ error("A") + trend("A") + season("A")),
    HWMult = ETS(Passengers ~ error("M") + trend("A") + season("M"))
  )

air_fc <- air_fit |>
  fabletools::forecast(h = 27)


fabletools::accuracy(air_fc, USAirlines_ts)

```
Based upone this information, which model would you choose?

## ETS

You can also allow the computer to search for the best model.  The ETS (Error, Trend, Seasonality) algorithm will search for the best model and estimate the parameters.  For the error term, we can have either an additive or multiplicative error structure.  For the trend, we can have none, additive, or damped additive . For the seasonal component, we can have none, additive or multiplicative (lots of choices!). An example of how to run this is:

```{r ets}
air_auto <- air_train |>
  model(ETS(Passengers))
report(air_auto)

# Now compare this to the HW models:

air_fit <- air_train |>
  model(
    HWAdd = ETS(Passengers ~ error("A") + trend("A") + season("A")),
    HWMult = ETS(Passengers ~ error("M") + trend("A") + season("M")),
    AutoETS = ETS(Passengers ~ error("M") + trend("Ad") + season("M"))
  )

air_fc <- air_fit |>
  fabletools::forecast(h = 27)


fabletools::accuracy(air_fc, USAirlines_ts)

### Can change to bic
air_auto2 <- air_train |>
  model(ETS(Passengers, ic ="bic"))
report(air_auto2)

```


## Python Code for Exponential Smoothing

The following Python codes will produce exponential smoothing models.  The exponential smoothing models are using an older version in statsmodels (a new format is in statsforecast).

```{python simple exponential smoothing}

import numpy as np
import pandas as pd
import matplotlib.pyplot as plt
from matplotlib import pyplot
from statsforecast.models import ETS
from statsforecast import StatsForecast
from statsforecast.utils import AirPassengers
from statsmodels.tsa.tsatools import detrend
from sklearn.metrics import mean_squared_error
from statsmodels.tsa.api import ExponentialSmoothing, SimpleExpSmoothing, Holt


steel=pd.read_csv("Q:\\My Drive\\Fall 2017 - Time Series\\DataR\\steel.csv")
df = pd.date_range(start='1932-07-01', end='1980-07-01', freq='6ME')
steel.index=pd.to_datetime(df)

steel_train=steel.head(87)
steel_test = steel.tail(9)

fit = SimpleExpSmoothing(steel_train['steelshp']).fit()
fit.params['smoothing_level']

fcast = fit.forecast(9)
fcast

plt.plot(steel_train["steelshp"],color="black")
plt.plot(fcast,color="blue")
plt.show()

```


The Holt models in Python:

```{python holt models}



fit1 = Holt(steel_train['steelshp']).fit()
fit1.summary()
fcast1 = fit1.forecast(9)
fcast1
fit2 = Holt(steel_train['steelshp'],exponential=True).fit()
fit2.summary()
fcast2 = fit2.forecast(9)
fcast2
fit3 = Holt(steel_train['steelshp'],damped=True).fit()
fit3.summary()
fcast3 = fit3.forecast(9)
fcast3
fit4 = Holt(steel_train['steelshp'],exponential=True,damped=True).fit()
fit4.summary()
fcast4 = fit4.forecast(9)
fcast4

ax=steel_train.plot(color="black",figsize=(12,8))
fcast1.plot(ax=ax,color="blue")
fcast2.plot(ax=ax,color="orange")
fcast3.plot(ax=ax,color="purple")
fcast4.plot(ax=ax,color="gray")
steel_test.plot(ax=ax,color="red")
plt.show()



```


Seasonal models in Python:

```{python seasonal models}

import pandas as pd
import matplotlib.pyplot as plt
from matplotlib import pyplot
from statsmodels.tsa.seasonal import seasonal_decompose
from statsmodels.tsa.seasonal import STL
usair_p=pd.read_csv("Q:\\My Drive\\Fall 2017 - Time Series\\DataR\\usairlines.csv")
df=pd.date_range(start='1/1/1990', end='3/1/2008', freq='MS')
usair_p.index=pd.to_datetime(df)


usair_train = usair_p.head(192)
usair_test = usair_p.tail(27)

#### Caution!!!!! These did not converge!!! Would NOT recommend using them!

usair_train['HWES_ADD'] = ExponentialSmoothing(usair_train['Passengers'],trend='add',seasonal='add',seasonal_periods=12).fit().fittedvalues

usair_train['HWES_MUL'] = ExponentialSmoothing(usair_train['Passengers'],trend='mul',seasonal='mul',seasonal_periods=12).fit().fittedvalues

usair_train[['Passengers','HWES_ADD','HWES_MUL']].plot(title='Holt Winters Exponential Smoothing: Additive and Multiplicative Seasonality')
plt.show()

fit1=ExponentialSmoothing(usair_train['Passengers'],trend='add',seasonal='mul',seasonal_periods=12).fit()
fit1.summary()

fit2=ExponentialSmoothing(usair_train['Passengers'],trend='add',seasonal='add',seasonal_periods=12).fit()
fit2.summary()

fcast1=fit1.forecast(27)
fcast2=fit2.forecast(27)

ax=usair_train["Passengers"].plot(color="black",figsize=(12,8))
fcast1.plot(ax=ax,color="blue")
fcast2.plot(ax=ax,color="orange")
usair_test["Passengers"].plot(ax=ax,color = "red")
plt.show()

error=usair_test['Passengers']-fcast1
MAE=np.mean(abs(error))
MAE
MAPE=np.mean(abs(error)/usair_test['Passengers'])
MAPE

```




## SAS Code for Exponential Smoothing Models

The following code is for Exponential Smoothing models in SAS.


_Create library for data sets_
libname Time 'Q:\My Drive\Fall 2017 - Time Series\Data';   
run;

_SIMPLE EXPONENTIAL SMOOTHING MODEL_

_Create a simple exponential smoothing model_   


proc esm data=Time.Steel print=all plot=all lead=24;   
	forecast steelshp / model=simple;   
run;   

_Create a simple exponential smoothing model with ID statement_ \  

proc esm data=Time.USAirlines print=all plot=all lead=24; \  
	id date interval=month; \  
	forecast Passengers / model=simple; \  
run; \  


_LINEAR TREND FOR EXPONENTIAL SMOOTHING_ \  

_Double exponential smoothing_ \  

proc esm data=Time.Steel print=all plot=all lead=24; \  
	forecast steelshp / model=double; \  
run; \  

_linear exponential smoothing_ \  

proc esm data=Time.Steel print=all plot=all lead=24; \  
	forecast steelshp / model=linear; \  
	run; \  
	

_damped trend exponential smoothing_ \  

proc esm data=Time.Steel print=all plot=all lead=24; \  
	forecast steelshp / model=damptrend; \  
	run; \  
	

_linear exponential smoothign with interval = month_ \  

proc esm data=Time.USAirlines print=all plot=all lead=24; \  
	id date interval=month; \  
		forecast Passengers / model=linear; \  
		run; \  
		

_SEASONAL EXPONENTIAL SMOOTHING MODEL_ \  

_Additive seasonal exponential smoothing model_ \  

proc esm data=Time.USAirlines print=all plot=all \  
		 seasonality=12 lead=24 outfor=test1; \  
		 forecast Passengers / model=addseasonal; \  
		 run; \  
		 
_mulitplicative seasonal exponential smoothing model_ \  

proc esm data=Time.USAirlines print=all plot=all \  
		 seasonality=12 lead=24; \  
		 	forecast Passengers / model=multseasonal; \  
		 	run; \  
		 	

_Winters additive exponential smoothing model (includes trend)_ \  

proc esm data=Time.USAirlines print=all plot=all \  
		 seasonality=12 lead=24; \  
		 forecast Passengers / model=addwinters; \  
		 run; \  
		 
_Winters multiplicative exponential smoothing model (includes trend) (Lead = 24)_ \  

proc esm data=Time.USAirlines print=all plot=all \  
		 seasonality=12 lead=24; \  
		 forecast Passengers / model=multwinters; \  
		 run; \  
		 
_Winters multiplicative exponential smoothing model (includes trend) Lead = 12_ \  

proc esm data=Time.USAirlines print=all plot=all lead=12 \  
  back=12 seasonality=12; \  
  forecast Passengers / model=multwinters; \  
  run; \  
  

_EXPLORATION of SEASONAL EXPONENTIAL SMOOTHING MODEL_ \  


_Winters multiplicative exponential smoothing model (includes trend) Lead = 12, uses outfor statement to output forecasts_ \  

proc esm data=Time.USAirlines print=all plot=all \
		 seasonality=12 lead=12 back=12 outfor=test; \  
		 	forecast Passengers / model=multwinters; \  
		 	run; \ 

_calculate |error|/|actual value|_ \  

data test2; \  
set test; \  
if _TIMEID_>207; \  
abs_error=abs(error); \  
abs_err_obs=abs_error/abs(actual); \  
run; \  


_mean of |error|/|actual value| for this forecast_ \  

proc means data=test2 mean; \  
var abs_error abs_err_obs; \  
run;



